Un Cadre Basé sur la Discrépance pour une Comparaison de la Robustesse entre Evaluations Multi-attributs

### 2016-12-22 Docs/translation/paper_FR.tex
more_tr

@@

A second assumption we make is that the multi-objective evaluation is done through linear aggregation of objectives, i.e. that we are tackling a multi-attribute optimization problem. If $(q_i(\vec{x}))_i$ are values of objectives functions, then weights $(w_i)_i$ are defined in order to build the aggregated decision-making function $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, which value determines then the performance of the solution. It is analog to aggregated utility techniques in economics and is used in many fields. The subtlety lies in the choice of weights, i.e. the shape of the projection function, and various approaches have been developed to find weights depending on the nature of the problem. Recent work~\cite{dobbie2013robustness} proposed to compare robustness of different aggregation techniques through sensitivity analysis, performed by Monte-Carlo simulations on synthetic data. Distribution of biases where obtained for various techniques and some showed to perform significantly better than others. Robustness assessment still depended on models used in that work.

@@

%%%%%%%%%%%%%%%%
%% Framework Description
%%%%%%%%%%%%%%%%
\section{Framework Description}

%%%%%%%%%%%%%%%%
\subsection{Intuitive Description}

We describe now the abstract framework allowing theoretically to compare robustnesses of evaluations of two different urban systems. Our framework is a generalization of an empirical method proposed in~\cite{ecodistrictReport} besides a more general benchmarking study on indicator sense and relevance in a sustainability context. Intuitively, it relies on empirical base resulting from the following axioms :
\begin{itemize}
\item Urban systems can be seen from the information available, i.e. raw data describing the system. As a data-driven approach, this raw data is the basis of our framework and robustness will be determined by its structure.
\item From data are computed indicators (objective functions). We assume that a choice of indicators is an intention to translate particular aspects of the system, i.e. to capture a realization of an ``urban fact'' (\emph{fait urbain}) in the sense of \noun{Mangin}~\cite{mangin1999projet} - a sort of stylized fact in terms of processes and mechanisms, having various realizations on spatially distinct systems, depending on each precise context.
\item Given many systems and associated indicators, a common space can be built to compare them. In that space, data represents more or less well real systems, depending e.g. on initial scale, precision of data, missing data. We precisely propose to capture that through the notion of point cloud discrepancy, which is a mathematical tool coming from sampling theory expressing how a dataset is distributed in the space it is embedded in~\cite{dick2010digital}. %Coupling discrepancies with appropriate weights depending on indicator importance allows to introduce a relative ratio of robustness between two evaluations.
\end{itemize}

Synthesizing these requirements, we propose a notion of \emph{Robustness} of an evaluation that captures both, by combining data reliability with relative importance,
\begin{enumerate}
\item \emph{Missing Data} : an evaluation based on more refined datasets will naturally be more robust.
\item \emph{Indicator importance} : indicators with more relative influence will weight more on the total robustness.
\end{enumerate}

%%%%%%%%%%%%%%%%
\subsection{Formal Description}

%%%%%%%%%%%%%%%%
\paragraph{Indicators}

Let $(S_{i})_{1\leq i\leq N}$ be a finite number of geographically disjoints territorial systems, that we assume described through raw data and intermediate indicators, yielding $S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}$ with $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$ such that each subspace contain real matrices : $\mathcal{X}_{i,k}=\mathbb{R}^{n_{i,k}^{X}p_{i,k}^{X}}$ (the same holding for $\mathcal{Y}_{i}$). We also define an ontological index function $I_{X}(i,k)$ (resp. $I_{Y}(i,k)$) taking integer values which coincide if and only if the two variables have the same ontology in the sense of~\cite{livet2010}, i.e. they are supposed to represent the same real object. We distinguish ``raw data'' $\mathbf{X}_{i}$ from which indicators are computed via explicit deterministic functions, from ``intermediate indicators'' $\mathbf{Y}_{i}$ that are already integrated and can be e.g. outputs of elaborated models simulating some aspects of the urban system. We define the partial characteristic space of the ``urban fact'' by

\begin{equation}
%\begin{split}
%(\mathcal{X},\mathcal{Y}) & \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right)\\
%& =  \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
%\end{split}
(\mathcal{X},\mathcal{Y}) \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right) = \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
\end{equation}

with $\mathcal{D}_{\mathcal{X}}=\{\mathcal{X}_{i,k}|I(i,k)\textrm{ distincts},n_{i,k}^{X}\mbox{ maximal}\}$
(the same holding for $\mathcal{Y}_{i}$). It is indeed the abstract space on which indicators are integrated. The indices $c$ introduced as a definition here correspond to different indicators across all systems. This space is the minimal space common to all systems allowing a common definition for indicators on each.

Let $\mathbf{X}_{i,c}$ be the data canonically projected in the corresponding subspace, well defined for all $i$ and all $c$. We make the key assumption that all indicators are computed by integration against a certain kernel, i.e. that for all $c$, there exists $H_{c}$ space of real-valued functions on $(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})$, such that for all $h\in H_{c}$ :
\begin{enumerate}
\item $h$ is ``enough'' regular (tempered distributions e.g.)
\item $q_c=\int_{(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})}h$ is a function describing the ``urban fact'' (the indicator in itself)
\end{enumerate}

Typical concrete example of kernels can be :

\begin{itemize}
\item A mean of rows of $\mathbf{X}_{i,c}$ is computed with $h(x)=x\cdot f_{i,c}(x)$ where $f_{i,c}$ is the density of the distribution of the assumed underlying variable.
\item A rate of elements respecting a given condition $C$, $h(x)=f_{i,c}(x)\chi_{C(x)}$ 
\item For already aggregated variables $\mathbf{Y}$, a Dirac distribution allows to express them also as a kernel integral. 
\end{itemize}

%%%%%%%%%%%%%%%%
\paragraph{Aggregation}

%As described in introduction, the multi-objective nature of the problem is tackled through linear aggregation. Although it may be seen as a reductive approach, it becomes an asset for the application of integral approximation theorems as the next step of our method. 
Weighting objectives in multi-attribute decision-making is indeed the crucial point of the processes, and numerous methods are available (see~\cite{wang2009review} for a review for the particular case of sustainable energy management). Let define weights for the linear aggregation. We assume the indicators normalized, i.e. $q_c \in [0,1]$, for a more simple construction of relative weights. For $i,c$ and $h_{c}\in H_{c}$ given, the weight $w_{i,c}$ is simply constituted by the relative importance of the indicator $w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}$ where $\hat{q}_{i,c}$ is an estimator of $q_{c}$ for data $\mathbf{X}_{i,c}$ (i.e. the effectively calculated value). Note that this step can be extended to any sets of weight attributions, by taking for example $\tilde{w}_{i,c} = w_{i,c} \cdot w'_{i,c}$ if $\mathbf{w}'$ are the weights attributed by the decision-maker. We focus here on the relative influence of attributes and thus choose this simple form for weights.

%  subjective part is shit.
% subjective weight : various potential methods (notations, importance order, etc.), reviewed in~\cite{wang2009review}, are available to give a normalized subjective weight. In practice, these will be set to zero in our first implementation as we test the framework on virtual cases.

%%%%%%%%%%%%%%%%
\paragraph{Robustness Estimation}

The scene is now set up to be able to estimate the robustness of the evaluation done through the aggregated function. Therefore, we apply an integral approximation method similar to methods introduced in~\cite{varet2010developpement}, since the integrated form of indicators indeed brings the benefits of such powerful theoretical results. Let $\mathbf{X}_{i,c}=(\vec{X}_{i,c,l})_{1\leq l\leq n_{i,c}}$ and $D_{i,c}=Disc_{\tilde{\mathcal{X}}_{c},L^2}(\mathbf{X}_{i,c})$ the discrepancy of data points cloud\footnote{The discrepancy is defined as the $L2$-norm of local discrepancy which is for normalized data points $\mathbf{X}=(x_{ij})\in \left[0,1\right]^d$, a function of $\mathbf{t}\in \left[0,1\right]^d$ comparing the number of points falling in the corresponding hypercube with its volume, by $disc(\mathbf{t}) = \frac{1}{n}\sum_i \mathbbm{1}_{\prod_j x_{ij}<t_j} - \prod_j t_j$. It is a measure of how the point cloud covers the space.}~\cite{niederreiter1972discrepancy}. With $h\in H_{c}$, we have the upper bound on the integral approximation error

@@

where $K$ is a constant independent of data points and objective function. It directly yields

@@

Assuming the error reasonably realized (``worst case'' scenario for knowledge of the theoretical value of aggregated function), we take this upper bound as an approximation of its magnitude. Furthermore, taking normalized indicators implies $\left|\left|\left|h_c\right|\right|\right| = 1$. We propose then to compare error bounds between two evaluations. They depend only on data distribution (equivalent to \emph{statistical robustness}) and on indicators chosen (sort of \emph{ontological robustness}, i.e. do the indicators have a real sense in the chosen context and do their values make sense), and are a way to combine these two type of robustnesses into a single value.

We thus define a \emph{robustness ratio} to compare the robustness of two evaluations by

@@

The intuitive sense of this definition is that one compares robustness of evaluations by comparing the highest error done in each based on data structure and relative importance.

By taking then an order relation on evaluations by comparing the position of the ratio to one, it is obvious that we obtain a complete order on all possible evaluations. This ratio should theoretically allow to compare any evaluation of an urban system. To keep an ontological sense to it, it should be used to compare disjoints sub-systems with a reasonable proportion of indicators in common, or the same sub-system with varying indicators. Note that it provides a way to test the influence of indicators on an evaluation by analyzing the sensitivity if the ratio to their removal. On the contrary, finding a ``minimal'' number of indicators each making the ratio strongly vary should be a way to isolate essential parameters ruling the sub-system.

%%%%%%%%%%%%%%%%
%% Results
%%%%%%%%%%%%%%%%
\section{Results}

%We apply our framework to a ``toy-model'', as elaborated indicator definition nor raw data collection was not the primary aim of this work, which was more to theoretically introduce the framework. It is however essential for such a data-driven paradigm to demonstrate through simulation the potentialities of the method, what we do here on semi-synthetic data.

%%%%%%%%%%%%%%%%
\paragraph{Implementation}

Preprocessing of geographical data is made through QGIS~\cite{qgis2011quantum} for performance reasons. Core implementation of the framework is done in R~\cite{team2000r} for the flexibility of data management and statistical computations. Furthermore, the package \texttt{DiceDesign}~\cite{franco20092} written for numerical experiments and sampling purposes, allows an efficient and direct computation of discrepancies. Last but not least, all source code is openly available on the \texttt{git} repository of the project\footnote{at \url{https://github.com/JusteRaimbault/RobustnessDiscrepancy}} for reproducibility purposes~\cite{ram2013git}.

%%%%%%%%%%%%%%%%
\subsection{Implementation on Synthetic Data}

We propose in a first time to illustrate the implementation with an application to synthetic data and indicators, for intra-urban quality indicators in the city of Paris.

%%%%%%%%%%%%%%%%
\paragraph{Data Collection}

We base our virtual case on real geographical data, in particular for \emph{arrondissements} of Paris. We use open data available through the OpenStreetMap project~\cite{bennett2010openstreetmap} that provides accurate high definition data for many urban features. We use the street network and position of buildings within the city of Paris. %\footnote{for which the third-party website \url{https://mapzen.com/metro-extracts/} provides already packaged extracts as for many metropolitan areas of the world}
Limits of \emph{arrondissements}, used to overlay and extract features when working on single districts, are also extracted from the same source. We use centroids of buildings polygons, and segments of street network. Dataset overall consists of around $200k$ building features and $100k$ road segments.
%An example of visualization of used data is shown in Fig.~\ref{fig:data_ex}.

%%%%%%%%%%%%%%%%
\paragraph{Virtual Cases}

We work on each district of Paris (from the 1st to the 20th) as an evaluated urban system. We construct random synthetic data associated to spatial features, so each district has to be evaluated many time to obtain mean statistical behavior of toy indicators and robustness ratios. The indicators chosen need to be computed on residential and street network spatial data. We implement two mean kernels and a conditional mean to show different examples, linked to environmental sustainability and quality of life, that are required to be maximized. Note that these indicators have a real meaning but no particular reason to be aggregated, they are chosen here for the convenience of the toy model and the generation of synthetic data. With $a\in \{1\ldots 20\}$ the number of the district, $A(a)$ corresponding spatial extent, $b\in B$ building coordinates and $s\in S$ street segments, we take
\begin{itemize}
\item Complementary of the average daily distance to work with car per individual, approximated by, with $n_{cars}(b)$ number of cars in the building (randomly generated by associated of cars to a number of building proportional to motorization rate $\alpha_m ~ 0.4$ in Paris), $d_w$ distance to work of individuals (generated from the building to a uniformly generated random point in spatial extent of the dataset), and $d_{max}$ the diameter of Paris area, $\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}$
\item Complementary of average car flows within the streets in the district, approximated by, with $\varphi(s)$ relative flow in street segment $s$, generated through the minimum of 1 and a log-normal distribution adjusted to have $95\%$ of mass smaller than 1 what mimics the hierarchical distribution of street use (corresponding to betweenness centrality), and $l(s)$ segment length, $\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}$

\item Relative length of pedestrian streets $\bar{p}$, computed through a randomly uniformly generated dummy variable adjusted to have a fixed global proportion of segments that are pedestrian.
\end{itemize}

@@

\caption{Numerical results of simulation for each district with $N=50$ repetitions. Each toy indicator value is given by mean on repetitions and associated standard deviation. Robustness ratio is computed relative to first district (arbitrary choice). A ratio smaller than 1 means that integral bound is smaller for upper district, i.e. that evaluation is more robust for this district. Because of the small size of first district, we expected a majority of district to give ratio smaller than 1, what is confirmed by results, even when adding standard deviations.}

@@

As synthetic data are stochastic, we run the computation for each district $N=50$ times, what was a reasonable compromise between statistical convergence and time required for computation. Table 1 shows results (mean and standard deviations) of indicator values and robustness ratio computation. Obtained standard deviation confirm that this number of repetitions give consistent results. Indicators obtained through a fixed ratio show small variability what may a limit of this toy approach. However, we obtain the interesting result that a majority of districts give more robust evaluations than 1st district, what was expected because of the size and content of this district : it is indeed a small one with large administrative buildings, what means less spatial elements and thus a less robust evaluation following our definition of the robustness.

%%%%%%%%%%%%%%%%
\subsection{Application to a Real Case : Metropolitan Segregation}

The first example was aimed to show potentialities of the method but was purely synthetic, hence yielding no concrete conclusion nor implications for policy. We propose now to apply it to real data for the example of metropolitan segregation.

\paragraph{Data}

We work on income data available for France at an intra-urban level (basic statistical units IRIS) for the year 2011 under the form of summary statistics (deciles if the area is populated enough to ensure anonymity), provided by INSEE\footnote{\texttt{http://www.insee.fr}}. Data are associated with geographical extent of statistical units, allowing computation of spatial analysis indicators.

\paragraph{Indicators}

>>>>

% encoding 
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}

@@

Une seconde hypothèse que nous faisons est que l'évaluation multi-objectifs est effectuée par agrégation linéaire des objectif, c'est à dire qu'on se place dans le cadre d'un problème d'optimisation multi-attributs. Si $(q_i(\vec{x}))_i$ sont les valeurs des fonctions objectifs, on définit alors des poids $(w_i)_i$ afin de construire la fonction de prise de décision $q(\vec{x})=\sum_i{w_i q_i(\vec{x})}$, dont la valeur détermine ensuite la performance d'une solution. Cette approche est analogue aux utilités agrégées en économie et est utilisée dans de nombreux domaines. La subtilité réside dans le choix des poids, i.e. de la forme de la fonction de projection, et différentes solutions ont été dévelopées pour obtenir des poids selon la nature du problème. Récemment, \cite{dobbie2013robustness} a proposé de comparer la robustesse des différentes techniques d'agrégation par une analyse de sensibilité, effectuée par simulations de Monte-Carlo pour produire des données synthétiques, ce qui permet d'obtenir la distribution des biais pour les différentes techniques, certaines étant significativement plus performantes que d'autres. Toutefois, la quantification de la robustesse dépend toujours des modèles utilisés dans ce travail.

@@

%%%%%%%%%%%%%%%%
%% Framework Description
%%%%%%%%%%%%%%%%
\section{Description du Cadre}

%%%%%%%%%%%%%%%%
\subsection{Description Intuitive}

Nous décrivons à présent le cadre proposé pour permettre théoriquement de comparer la robustesse d'évaluation de deux systèmes urbains différents. Ce cadre est une généralisation d'une méthode empirique proposée dans~\cite{ecodistrictReport} pour accompagner une étude dans un autre contexte effectuant une comparaison du sens et de la pertinence des indicateurs dans un contexte de durabilité. Intuitivement, la base empirique se base sur les principes suivants :

\begin{itemize}
\item Les systèmes urbains peuvent être vus selon l'information disponible, i.e. les données brutes décrivant le système. Dans une approche basée sur les données, celles-ci sont la base de notre cadre et la robustesse sera déterminée par leur structure.
\item A partir des données sont capturés des indicateurs (fonctions objectifs). Nous supposons qu'un choix d'indicateurs est une intention particulière de traduire des aspects particuliers du système, i.e. de capturer une réalisation d'un ``fait urbain'' au sens de \noun{Mangin}~\cite{mangin1999projet} - une sorte de fait stylisé en terme de processus et de mécanismes, ayant différentes réalisations sur des systèmes distincts dans l'espace, dépendant de chaque contexte géographique précis.
\item Etant donné plusieurs systèmes et indicateurs associés, un espace commun peut être construit pour les comparer. Dans cet espace, les données représentent plus ou moins bien le système réel, c'est à dire qu'elles sont imprécises en fonction de l'échelle initiale, de la précision effective des données. Nous proposons de capturer exactement ces différents aspects au travers de la notion de discrépance d'un nuage de points, qui est un outil mathématique provenant des théories d'échantillonnage, permettant d'exprimer la façon dont un jeu de données rempli l'espace dans lequel il s'insère~\cite{dick2010digital}.
\end{itemize}

Synthétisant ces contraintes, nous proposons une notion de \emph{Robustesse} d'une évaluation qui capture à la fois, en combinant la fiabilité des données à l'importance relative des indicateurs,

\begin{enumerate}
\item \emph{Données manquantes} : une évaluation se basant sur des jeux de données plus raffinés sera naturellement plus robuste.
\item \emph{Importance des indicateurs} : les indicateurs avec plus d'importance relative pèseront plus dans la robustesse totale.
\end{enumerate}

%%%%%%%%%%%%%%%%
\subsection{Description Formelle}

%%%%%%%%%%%%%%%%
\paragraph{Indicateurs}

Soit $(S_{i})_{1\leq i\leq N}$ un nombre fini de systèmes territoriaux géographiquement disjoints, % TODO : Q pourquoi nécessaire des les avoir spatially disjoints, could be different indicators on the same area ? maybe makes less sense ? missing point for comparability ?
que nous supposons décrits par les données brutes et des indicateurs intermédiaires, donnés par $S_{i}=(\mathbf{X}_{i},\mathbf{Y}_{i})\in\mathcal{X}_{i}\times\mathcal{Y}_{i}$ avec $\mathcal{X}_{i}=\prod_{k}\mathcal{X}_{i,k}$ tel que chaque sous-espace contient des matrices réelles : $\mathcal{X}_{i,k}=\mathbb{R}^{n_{i,k}^{X}p_{i,k}^{X}}$ (de la même façon pour $\mathcal{Y}_{i}$). Nous définissons également une fonction d'indice ontologique $I_{X}(i,k)$ (resp. $I_{Y}(i,k)$) prenant des valeurs entières qui coincident si et seulement si les deux variables ont même ontologie au sens de~\cite{livet2010}, c'est à dire qu'elles sont supposées représenter le même objet réel. On distingue les ``données brutes'' $\mathbf{X}_{i}$ à partir desquelles les indicateurs sont calculés généralement par des fonctions déterministes explicites, % TODO not that free on the computation here !
 des ``indicateurs intermédiaires'' $\mathbf{Y}_{i}$ qui sont déjà intégrés et peuvent être par exemple les sorties de modèles élaborés simulant certains aspects du système urbain. Nous définissons l'espace caractéristique du ``fait urbain'' par

\begin{equation}
(\mathcal{X},\mathcal{Y}) \underset{def}{=} \left(\prod\tilde{\mathcal{X}}_{c}\right)\times\left(\prod\tilde{\mathcal{Y}}_{c}\right) = \left(\prod_{\mathcal{X}_{i,k}\in\mathcal{D}_{\mathcal{X}}}\mathbb{R}^{p_{i,k}^{X}}\right)\times\left(\prod_{\mathcal{Y}_{i,k}\in\mathcal{D}_{\mathcal{Y}}}\mathbb{R}^{p_{i,k}^{Y}}\right)
\end{equation}

avec $\mathcal{D}_{\mathcal{X}}=\{\mathcal{X}_{i,k}|I(i,k)\textrm{ distincts},n_{i,k}^{X}\mbox{ maximal}\}$
(de même pour $\mathcal{Y}_{i}$). Il s'agit en fait de l'espace abstrait sur lequel les indicateurs sont intégrés. Les indices $c$ introduit par définition correspondent aux différents indicateurs au sein des différents systèmes. Cette espace est l'espace minimal commun à tous les systèmes permettant une définition commune des indicateurs pour tous.

Soit $\mathbf{X}_{i,c}$ les données projetées canoniquement sur le sous-espace correspondant, bien définies pour tout $i$ et tout $c$. Nous faisons donc l'hypothèse clé que tous les indicateurs sont calculés par intégration contre un noyau donné, i.e. pour tout $c$ il existe $H_{c}$ espace de fonctions à valeurs réelles sur $(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})$, tel que pour tout $h\in H_{c}$ :

\begin{enumerate}
\item $h$ est ``suffisamment'' régulière (distribution tempérée par exemple)
\item $q_c=\int_{(\tilde{\mathcal{X}}_{c},\tilde{\mathcal{Y}}_{c})}h$ est une fonction décrivant le ``fait urbain'' (l'indicateur en lui-même)
\end{enumerate}

Des exemples typiques de noyaux peuvent être :

\begin{itemize}
\item Une moyenne des lignes de $\mathbf{X}_{i,c}$ est calculée par $h(x)=x\cdot f_{i,c}(x)$ où $f_{i,c}$ est la densité de la distribution de la variable sous-jacente.
\item Un taux d'éléments du jeu de données respectant une condition donnée $C$, $h(x)=f_{i,c}(x)\chi_{C(x)}$.
\item Pour des variables déjà agrégées $\mathbf{Y}$, une distribution de Dirac permet des les exprimer également comme des intégrales de noyaux.
\end{itemize}

%%%%%%%%%%%%%%%%
\paragraph{Agrégation}

La détermination des poids est en fait le point crucial des processus de prise de décision multi-attributs, et de nombreuses méthodes sont disponibles (voir~\cite{wang2009review} pour une revue dans le cas particulier de la gestion de l'énergie durable). Définissons les poids pour l'agrégation linéaire. Nous supposons les indicateurs normalisés, i.e. $q_c \in [0,1]$, pour une construction plus simple des poids relatifs. % TODO : indeed $h_c \in [0,1]$ is the right assumption
Pour $i,c$ et $h_{c}\in H_{c}$ donnés, le poids $w_{i,c}$ est simplement constitué par l'importance relative de l'indicateur $w_{i,c}^{L}=\frac{\hat{q}_{i,c}}{\sum_{c}\hat{q}_{i,c}}$ où $\hat{q}_{i,c}$ est un estimateur de $q_{c}$ pour les données $\mathbf{X}_{i,c}$ (i.e. la valeur calculée effectivement). On peut noter que cette étape n'est pas contraignante et que cela peut être étendu à tout ensemble d'attribution de poids, en prenant par exemple $\tilde{w}_{i,c} = w_{i,c} \cdot w'_{i,c}$ si $\mathbf{w}'$ sont les poids fixés par le preneur de décisions. Nous nous concentrons sur l'influence relative des attributs et pour cela choisissons cette forme simple pour les poids.

%%%%%%%%%%%%%%%%
\paragraph{Estimation de la Robustesse}

La scène est à présent prête pour permettre d'estimer la robustesse d'une évaluation faite par la fonction d'agrégation. Pour cela, nous appliquons un théorème d'approximation d'intégrale similaire au méthodes introduites dans~\cite{varet2010developpement}, puisque la forme intégrée des indicateurs permet justement de bénéficier de tels résultats théoriquement puissant. Soit $\mathbf{X}_{i,c}=(\vec{X}_{i,c,l})_{1\leq l\leq n_{i,c}}$ et $D_{i,c}=Disc_{\tilde{\mathcal{X}}_{c},L^2}(\mathbf{X}_{i,c})$ le discrépance du jeu de données\footnote{La discrépance est définie comme la norme-$L2$ de la discrépance locale qui est pour des points de données normalisés $\mathbf{X}=(x_{ij})\in \left[0,1\right]^d$, une fonction de $\mathbf{t}\in \left[0,1\right]^d$ comparant le nombre de points compris dans le volume de l'hypercube correspondant, donné par $disc(\mathbf{t}) = \frac{1}{n}\sum_i \mathbbm{1}_{\prod_j x_{ij}<t_j} - \prod_j t_j$. C'est une mesure de la manière dont le nuage de points couvre l'espace.}~\cite{niederreiter1972discrepancy}. Avec $h\in H_{c}$, on a la borne supérieure sur l'erreur d'approximation de l'intégrale

@@

où $K$ est une constante indépendante des points de données et des fonctions objectifs. Cela donne directement

@@

En supposant l'erreur réalisée de manière raisonnable (scénario du ``pire de cas'' pour la connaissance de la valeur théorique de la fonction agrégée), nous prenons cette borne supérieure comme une approximation de sa magnitude. De plus, la normalisation des indicateurs implique que $\left|\left|\left|h_c\right|\right|\right| = 1$. Nous proposons alors de comparer les bornes d'erreurs entre deux évaluations. Elle dépendent seulement de la distribution des données (équivalence à la \emph{robustesse statistique}) et des indicateurs choisis (sorte de \emph{robustesse ontologique}, i.e. est-ce que les indicateurs ont un sens réel dans le contexte choisi et est-ce que leur valeur fait sens), et sont un moyen de combiner ces deux types de robustesse dans une seule valeur.

Nous définissons ainsi un \emph{ratio de robustesse} pour comparer la robustesse de deux évaluations par

@@

L'interprétation intuitive de cette définition est que l'on compare la robustesse des évaluations en comparant la plus grande erreur faite dans chaque cas selon la structure des données et l'importance relative.

En construisant une relation d'ordre sur les évaluations en comparant la position du ratio par rapport à un, il est clair qu'on obtient un ordre complet sur l'ensemble des évaluations possibles. Ce ratio devrait en théorie permettre de comparer n'importe quelle évaluation d'un système urbain. Afin de garder un sens ontologique à cela, il devrait être utilisé pour comparer des sous-systèmes disjoints  avec une proportion raisonnable d'indicateurs en commun, ou le même sous-système avec des indicateurs différents. On peut noter que cela fournit un moyen de tester l'influence des indicateurs sur une évaluation, en analysant la sensibilité du ratio à leur suppression. Au contraire, la détermination d'un nombre ``minimal'' d'indicateurs faisant chacun varier le ratio fortement pourrait être un moyen d'isoler des paramètres essentiels régissant le sous-système.

%%%%%%%%%%%%%%%%
%% Results
%%%%%%%%%%%%%%%%
\section{Résultats}

%%%%%%%%%%%%%%%%
\paragraph{Implémentation}

Le pré-traitement des données géographiques est fait via QGIS~\cite{qgis2011quantum} pour des raisons de performances. % TODO plutôt ergonomie ?
L'implémentation du coeur est faite en R~\cite{team2000r} pour la flexibilité de la gestion des données et du traitement statistique. De plus, le package \texttt{DiceDesign}~\cite{franco20092} conçu pour les expériences numériques et l'échantillonnage, permet un calcul efficient et direct des discrépances. Enfin, tout aussi important, l'ensemble du code source est disponible de manière ouverte sur le dépôt \texttt{git}du projet\footnote{à \url{https://github.com/JusteRaimbault/RobustnessDiscrepancy}} pour permettre la reproductibilité et la réutilisation~\cite{ram2013git}.

%%%%%%%%%%%%%%%%
\subsection{Implémentation sur des Données Synthétiques}

Nous proposons dans un premier temps d'illustrer l'implémentation par une application à des données et indicateurs synthétiques, pour des indicateurs de qualité de vie intra-urbaine pour la ville de Paris.

%%%%%%%%%%%%%%%%
\paragraph{Données Brutes}

Le cas virtuel se base sur des données géographiques réelle, en particulier pour les arrondissements parisiens. Nous utilisons les données disponibles par le projet OpenStreetMap~\cite{bennett2010openstreetmap} qui fournit déjà des données précises en haute définition pour de nombreux aspects urbains. Nous utilisons le réseau de rues et la position des bâtiments dans la ville de Paris. Les limites des arrondissements, utilisées pour agréger et extraire les features lorsqu'on travaille sur un seul district, sont aussi pris de la même source. Nous utilisons les centroïdes des polygones des bâtiments et les segments du réseau de rues. Le jeu de données brutes consiste d'environ $200k$ bâtiments et $100k$ segments de rues.

%%%%%%%%%%%%%%%%
\paragraph{Indicateurs synthétiques}

Nous travaillons sur chaque arrondissement de Paris (du 1er au 20ème) comme un système urbain évalué. Des données synthétiques aléatoires sont associées aux features spatiales, chaque arrondissement pouvant alors être évalué de manière stochastique, et des répétitions permettent d'obtenir le comportement statistique moyen des indicateurs jouets et des ratios de robustesse. Les indicateurs choisis doivent être calculés comme des indicateurs résidentiels et du réseau de rues. Pour montrer différents exemples, nous implémentons deux kernels moyens et une moyenne conditionnelle, tous liés à la durabilité environnementale et la qualité de vie, chacun devant être maximisés. On peut noter que ces indicateurs ont un sens réel mais pas de raison particulière d'être agrégés, ils sont ici choisis pour l'aspect pratique du modèle jouet et de la génération de données synthétiques. Avec $a\in \{1\ldots 20\}$ le nombre d'arrondissements, $A(a)$ l'aire spatiale correspondante à chacun, $b\in B$ les coordonnées des bâtiments et $s\in S$ les segments de rues, nous prenons

\begin{itemize}
\item Le complémentaire de la distance journalière moyenne au travail en voiture par individu, approché par, avec $n_{cars}(b)$ nombre de voiture dans le bâtiment (généré aléatoirement en associant des voiture à bâtiments proportionnel au taux de motorisation attendu $\alpha_m ~ 0.4$ à Paris), $d_w$ distance des individus à leur travail (généré à partir du bâtiment vers un point aléatoire distribué uniformément dans l'étendue spatiale du jeu de données), et $d_{max}$ le diamètre de l'aire de Paris, $\bar{d}_w = 1 - \frac{1}{|b\in A(a)|} \cdot \sum_{b\in A(a)}{n_{cars}(b)\cdot \frac{d_w}{d_{max}}}$
\item Le complémentaire des flots moyens de voitures des rues dans la zone, approché par, avec $\varphi(s)$ flot relatif dans le segment de rue $s$, généré par le minimum entre 1 et une distribution log-normale ajustée pour avoir $95\%$ de masse plus petite que 1, ce qui mimique la distribution hiérarchique de l'utilisation des rues (qui correspond à la centralité de chemin), et $l(s)$ longueur du segment, $\bar{\varphi} = 1 - \frac{1}{|s\in A(a)|} \cdot \sum_{s \in A(a)}{\varphi(s)\cdot \frac{l(s)}{\max{(l(s))}}}$
\item Longueur relative de rues piétonnes $\bar{p}$, calculé vie une dummy variable aléatoire uniforme ajustée pour obtenir une proportion fixée de segments pédestre.
\end{itemize}

@@

\caption{Résultats numériques des simulations pour chaque arrondissement avec $N=50$ répétitions. Chaque valeur des indicateurs factice est donnée par sa moyenne sur les répétitions et la déviation standard associée. Le ratio de robustesse est calculé par rapport au premier arrondissement (choix arbitraire). Un ratio inférieur à 1 signifie que la borne de l'intégrale est plus petite pour le premier système, i.e. que l'évaluation est plus robuste pour celui-ci. % TODO : wrong à l'oral 15th block size ?
 A cause de la petite taille du 1er arrondissement, on s'attend que la majorité des arrondissements aient un ratio plus petit que 1, ce qui se confirme dans les résultats même lorsque l'on ajoute la déviation standard.}

@@

Comme les données synthétiques sont stochastiques, les simulations sont lancées pour chaque quartier $N=50$ fois, ce qui était un compromis raisonnable entre convergence statistique et temps nécessaire au calcul. La table 1 montre les résultats (moyennes et déviations standard) des valeurs des indicateurs et le calcul du ratio de robustesse. Les déviations standard obtenues confirment que ce nombre de simulations donnent des résultats consistants. Les indicateurs obtenus en fixant un ratio fixe montre peu de variabilité, ce qui peut être une limite de cette approche jouet. On obtient toutefois le résultat intéressant que la majorité des arrondissements donne des évaluations plus robustes que le 1er arrondissement, ce qui était attendu par la taille et la fonction de ce quartier: il s'agit en effet d'un petit quartier avec de grand bâtiment administratifs, ce qui implique moins d'éléments spatiaux et pour cela une évaluation moins robuste selon la définition qu'on en a donnée.

% TODO : question : check the claim on block size, maybe it's only because bigger ?
%   -- -> cf up --

%%%%%%%%%%%%%%%%
\subsection{Application à un Cas Réel: Ségrégation Métropolitaine}

Le premier exemple avait pour but de montrer les potentialités de la méthode mais était purement synthétique, ne pouvant pour cela fournir pas de conclusion concrete ni d'implications pour la gouvernance. Nous proposons maintenant de l'appliquer à des données réelles dans le cas de la ségrégation métropolitaine.

\paragraph{Données}

Nous travaillons sur les données de revenus, disponible pour la France à un niveau intra-urbain (unités statistiques élémentaires IRIS) pour l'année 2011 sous la forme de résumé statistiques (déciles uniquement si la zone est peuplée suffisamment pour assurer l'anonymat), fournies par l'INSEE\footnote{\texttt{http://www.insee.fr}}. Les données sont associées à l'étendue géographique des unités statistiques, permettant le calcul d'indicateurs d'analyse spatiale.

\paragraph{Indicateurs}
